---
title: "Intermediate Biostatistics Project"
output: 
  prettydoc::html_pretty: 
    fig_height: 6
    fig_width: 9
    highlight: github
    theme: tactile
    toc: yes
    df_print: kable
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load required library

First lets start with loading tidyverse library.

```{r  message=FALSE}
library(tidyverse)
```


# Data Wrangling

convert column to appropriate type and read the file

```{r message=FALSE, warning=FALSE}
types <- 'idddddddii'
fram<- read_csv("framingham_200.csv", col_types = types)
```


select the columns that we need

```{r message=FALSE, warning=FALSE}
fram <- select(fram, TOTCHOL, AGE,
               DIABP,CIGPDAY,BMI,HEARTRTE,
               GLUCOSE,STROKE,CVD)
```


convert to tibble format

```{r}
fram <- as_tibble(fram)
```

## Check missing Values

now see whether there are missing value in the **fram** data set

```{r  message=FALSE}
library(Amelia)
```

create missing map: in the map the yellow represents missing value

```{r echo=TRUE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE, fig.align='center'}
missmap(fram, legend = T, col=c('yellow','black'), main = 'Missing value map')
```

Now drop the missing value, following code will drop all the rows that contain missing value.

```{r}
fram <- fram %>% drop_na()
```

 now check the missing value again.
as you can see there are no more missing value in our data set

```{r fig.height=4, fig.width=6, fig.align='center'}
missmap(fram, legend = T, col=c('yellow','black'), main = 'Missing value map')
```

we don't need package **Amelia** anymore, so we will detach it

```{r}
detach('package:Amelia', unload = TRUE)
```


take a look to our data set

```{r}
glimpse(fram)
```

now convert stroke and cvd from integer to as factor

```{r}
fram$STROKE = as.factor(fram$STROKE)
fram$CVD = as.factor(fram$CVD)
```

take a look again to our data set

```{r}
glimpse(fram)
```

# Exploratory Data Analysis

## Plot

### QQPLOT

Now check the linearty plotting Q-Q plot 

```{r  fig.align='center'}

par(mfrow=c(3,3))
q1<- qqnorm(fram$TOTCHOL, main = 'Normal Q-Q Plot for TOTCHOL')
qqline(fram$TOTCHOL, col = "red")

q2<- qqnorm(fram$AGE, main = 'Normal Q-Q Plot for AGE')
qqline(fram$AGE, col = "red")

q3 <- qqnorm(fram$DIABP, main = 'Normal Q-Q Plot for DIABP')
qqline(fram$DIABP, col = "red")

q4<- qqnorm(fram$CIGPDAY, main = 'Normal Q-Q Plot for CIGPDAY')
qqline(fram$CIGPDAY, col = "red")

q5<- qqnorm(fram$BMI, main = 'Normal Q-Q Plot for BMI')
qqline(fram$BMI, col = "red")


q6<- qqnorm(fram$HEARTRTE, main = 'Normal Q-Q Plot for HEARTRTE')
qqline(fram$HEARTRTE, col = "red")

q7 <- qqnorm(fram$GLUCOSE, main = 'Normal Q-Q Plot for GLUCOSE')
qqline(fram$GLUCOSE, col = "red")

```


### QQPLOT in ggplot2




```{r fig.height=4, fig.width=6, fig.align='center'}


ggplot(fram) +
  stat_qq(aes(sample = TOTCHOL))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for Serum Total Cholesterol')+
  theme(plot.title = element_text(hjust = 0.5))



ggplot(fram) +
  stat_qq(aes(sample = AGE))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for Age')+
  theme(plot.title = element_text(hjust = 0.5))


ggplot(fram) +
  stat_qq(aes(sample = DIABP))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for DIABP')+
  theme(plot.title = element_text(hjust = 0.5))



ggplot(fram) +
  stat_qq(aes(sample = CIGPDAY))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for CIGPDAY')+
  theme(plot.title = element_text(hjust = 0.5))

ggplot(fram) +
  stat_qq(aes(sample = BMI))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for BMI')+
  theme(plot.title = element_text(hjust = 0.5))


ggplot(fram) +
  stat_qq(aes(sample = HEARTRTE))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for Heart Rate')+
  theme(plot.title = element_text(hjust = 0.5))

ggplot(fram) +
  stat_qq(aes(sample = GLUCOSE))+
  xlab('Theoretical Quantiles')+ ylab('Sample Quantiles')+ 
  ggtitle('Q-Q plot for Glucose')+
  theme(plot.title = element_text(hjust = 0.5))


```

load RcolorBrewer package as we will need it 

```{r message=FALSE, warning=FALSE}
library(RColorBrewer)
```

### Boxplot

Now we will check outlier using boxplot

this one for only Total cholesterol
```{r fig.height=4, fig.width=6, fig.align='center'}
boxplot( fram$TOTCHOL,  col = brewer.pal(1, "Pastel2" ), boxwex=.4, xlab= 'TOTCHOL')
```

this one for other continuous variable

```{r fig.height=4, fig.width=6, fig.align='center'}
boxplot( fram$AGE, fram$DIABP, 
         fram$CIGPDAY, fram$BMI,
         fram$HEARTRTE, fram$GLUCOSE,
         col = brewer.pal(6, "Pastel2" ), boxwex=.4,
         names =c('AGE','DIABP','CIGPDAY',
                  'BMI','HEARTRTE','GLUCOSE'))
```

now remove the stroke and cvd column  from **fram** because those are categorical valueand and save in **fram_hist** for drawing histogram and finding *skewness* and *kurtosis* value. we will also do shapiro-wilk test.


```{r}
fram_hist <- select(fram, -STROKE, -CVD)
```

##  skewness and kurtosis value

```{r message=FALSE, warning=FALSE}
library(moments)
```

```{r}
lapply(fram_hist, skewness)
lapply(fram_hist, kurtosis)
```

now do shapiro-wilk test
```{r}
lapply(fram_hist, shapiro.test)
```

```{r}
detach('package:moments', unload = TRUE)
```

lets put together all thing in one slide.
first load **psych** package

```{r message=FALSE, warning=FALSE}
library(psych)
```

```{r}
names<- c('TOTCHOL: SK-0.405: ', 'AGE: SK-0.177 ','DIABP: SK- 0.42',
          'CIGPDAY: SK-1.66','BMI: SK-0.84','HEARTRTE: SK-1.37','GLUCOSE: SK-1.64')


colnames(fram_hist)<-
  c('TOTCHOL \nSkewness-0.405 : Kurtosis:3.09: Shapiro-wilk test: 0.0058', 
    'AGE\nSkewness:0.177 : Kurtosis:2.46: p-value: 0.099',
    'DIABP\nSkewness:0.42: Kurtosis:3.14: p-value:0.00668' ,
    'CIGPDAY\nSkewness:1.66: Kurtosis:5.07: p-value: 2.2e-16' ,
    'BMI\nSkewness:0.84: Kurtosis:4.91: p-value:1.908e-06' ,
    'HEARTRTE\nSkewness:1.37: Kurtosis:6.77: p-value:2.173e-10 ',
    'GLUCOSE\nSkewness:1.64: Kurtosis:8.54: p-value:1.107e-11 ')



fh <- as.matrix(fram_hist) # convert to matrix

```

```{r fig.align='center'}
multi.hist(fh, bcol = 'skyblue', dcol = "red") 
```


```{r}
detach('package:psych', unload = TRUE)
```

## Correlation Matrix

load few more package for correlation matrix

```{r message=FALSE, warning=FALSE}
library(viridis)
library(corrplot)
library(corrgram)
```



```{r fig.height=4, fig.width=6, fig.align='center'}

fram_corr <- select(fram, -STROKE, -CVD)

cor.data <- cor(fram_corr)

corrplot(cor.data, method = "number", col = brewer.pal(8,'Dark2')) #first one

corrplot(cor.data, method = "color") #second one 

corrgram(fram_corr, order = TRUE, 
         lower.panel = panel.shade, 
         upper.panel = panel.pie, 
         text.panel = panel.txt) #third one

```

```{r}
detach('package:viridis', unload = TRUE)
detach('package:corrplot', unload = TRUE)
detach('package:corrgram', unload = TRUE)
```

we can do correlation matrix few other way

```{r fig.align='center'}
pairs(  ~  TOTCHOL+
        AGE +
        DIABP+
        BMI+
        HEARTRTE+
        GLUCOSE ,data = fram, main= 'Scatterplot Matrix')
```


```{r fig.align='center', fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
library(GGally)

#fram_corr is data set without stroke and cvd column

ggpairs(fram_corr)


```

```{r}
detach('package:GGally', unload = TRUE)
```


 ### correlation test for each variable with p value
 
```{r message=FALSE, warning=FALSE}
library(Hmisc)
```
 
```{r}
rcorr(as.matrix(fram_corr))
```

```{r}
detach('package:Hmisc', unload = TRUE)
```

# Modeling

 multiple linear regression with all independent variable

```{r}
model_all <- lm(TOTCHOL ~ 
                  AGE +
                  DIABP+
                  CIGPDAY+
                  BMI+
                  HEARTRTE+
                  GLUCOSE+
                  STROKE+
                  CVD,  data = fram) # you can also use `lm(TOTCHOL~ . , data=fram)`
```

```{r}
summary(model_all)
anova(model_all)
```

## Diagnostics

for diagnostic plot
```{r fig.height=4, fig.width=6, fig.align='center'}
plot(model_all)
```



few other command that might be useful <br>

* coefficients(model_all) # model coefficients  
* confint(model_all, level=0.95) # CIs for model parameters   
* fitted(model_all) # predicted values 
* residuals(model_all) # residuals  
* anova(model_all) # anova table   
* vcov(model_all) # covariance matrix for model parameters  
* influence(model_all) # regression diagnostics 


since only age and diabp are significant, lets do model with only those two independent variable

```{r}
model_age_diabp<- lm(TOTCHOL~
                       AGE+
                       DIABP, data = fram)
summary(model_age_diabp)
anova(model_age_diabp)
```

for diagnostic plot 

```{r fig.height=4, fig.width=6, fig.align='center'}
plot(model_age_diabp)
```

## Log Transformation

log transformation of variable and save in column

```{r}
fram <- fram %>% as_tibble() %>% 
  mutate(TOTCHOL_log = log10(TOTCHOL),
         DIABP_log = log10(DIABP),
         
         CIGPDAY_log = log10(CIGPDAY+1),
         BMI_log = log10(BMI),
         
         HEARTRTE_log= log10(HEARTRTE),
         GLUCOSE_log = log10(GLUCOSE))
```

Q-Q plot again

```{r fig.align='center'}

par(mfrow=c(2,3))
q1<- qqnorm(fram$TOTCHOL_log, 
            main="QQ-plot for total cholesterol")
qqline(fram$TOTCHOL_log, col = "red")


q2<- qqnorm(fram$BMI_log, main= 'QQ-plot for BMI')
qqline(fram$BMI_log, col = "red")

q3<- qqnorm(fram$DIABP_log, main = 'QQ-plot for DIABP')
qqline(fram$DIABP_log, col = "red")

q4<- qqnorm(fram$HEARTRTE_log, main = 'QQ-plot for HEART RATE')
qqline(fram$HEARTRTE_log, col = "red")

q5<- qqnorm(fram$GLUCOSE_log, main = 'QQ-plot for Glucose')
qqline(fram$GLUCOSE_log, col = "red")

q6<- qqnorm(fram$CIGPDAY_log, main = 'QQ-plot for CIGPDAY')
qqline(fram$CIGPDAY_log, col = "red")

```

### model with log transformed data
multiple linear regression with all independent variable

```{r}
model_all_log <- lm(TOTCHOL ~ 
                  AGE +
                  DIABP_log+
                  CIGPDAY+
                  BMI_log+
                  HEARTRTE_log+
                  GLUCOSE_log+
                  STROKE+
                  CVD,  data = fram)

summary(model_all_log)
anova(model_all_log)
```


```{r}
model_age_diabp_log <- lm(TOTCHOL ~ 
                            AGE +
                            DIABP_log, 
                          data = fram)

summary(model_age_diabp_log)
anova(model_age_diabp_log)
```

### multicolinearity

```{r message=FALSE, warning=FALSE}
library(car)
```

variance inflation factor(VIF)

It quantifies the severity of multicollinearity in an ordinary least squares regression analysis.
value less than 5 is acceptable and it (<5) means no significant multicolinearity between independent variables.


```{r}
vif(model_all)
vif(model_all_log)
vif(model_age_diabp)
vif(model_age_diabp_log)
```

```{r}
detach('package:car', unload = TRUE)
```


## model selection by AIC 

```{r message=FALSE, warning=FALSE}
library(MASS)
```

```{r}
stepAIC(model_all_log, direction = 'backward')
```

```{r message=FALSE, warning=FALSE}
detach('package:MASS', unload = TRUE)
```

# final model

$$\hat{y}= -81.4884 + 0.8553 * AGE + 143.7133* log_{10}(DIABP)$$


### The End